---
title: "GASPARONI dataset"
author: "Lanyu Zhang, Tiago C. Silva, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: lumen
    highlight: null
    toc: true
    number_sections: true
    df_print: paged
    code_download: true
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data retrival

```{R, message = FALSE, results = 'hide'}
library(dplyr)
library(minfi)
```

```{R}
cohort <- "GASPARONI"
data.dir <- file.path("DATASETS/",cohort,"/") 
data.dir.table <- "DATASETS/Summary_Table/" 
data.dir.raw <- file.path(data.dir,"/step1_download/") 
data.dir.read <- file.path(data.dir,"/step2_read_minfi/") 
data.dir.bsfilter <- file.path(data.dir,"/step3_bsConvFilter/") 
data.dir.clinical.filter <- file.path(data.dir,"/step4_clinical_available_filtering/") 
data.dir.probes.qc <- file.path(data.dir,"/step5_probesQC_filtering/") 
data.dir.probes.normalization <- file.path(data.dir,"/step6_normalization/") 
data.dir.pca <- file.path(data.dir,"/step7_pca_filtering/") 
data.dir.neuron <- file.path(data.dir,"/step8_neuron_comp/") 
data.dir.single.cpg.pval <- file.path(data.dir,"/step9_single_cpg_pval/") 
data.dir.residuals <- file.path(data.dir,"/step10_residuals/") 
data.dir.median <- file.path(data.dir,"/step11_median/") 
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

Required R libraries: `GEOquery` can be installed as following:

```{R, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("GEOquery")
```

Data from: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE66351

```{r synapse, eval = FALSE}
library(GEOquery)
library(SummarizedExperiment)
test <- getGEO(GEO = "GSE66351",destdir = data.dir.raw)
se <- test$GSE66351_series_matrix.txt.gz %>% makeSummarizedExperimentFromExpressionSet()
```

```{R, eval = FALSE}
se <- se[,se$source_name_ch1  == "Frontal Cortex" &
           se$characteristics_ch1 == "cell type: bulk" &
           se$braak_stage.ch1 != "NA"]
```

We  will download Idat files for all our samples

```{R, eval =  FALSE}
files <- c(se$supplementary_file %>% as.character(),
           se$supplementary_file.1 %>% as.character())
dir.create(data.dir.raw,showWarnings = FALSE,recursive = TRUE)
plyr::a_ply(files,.margins = 1,.fun = function(url){
  out.file <- file.path(data.dir.raw,basename(url))
  if(!file.exists(out.file)) downloader::download(url,out.file)
  gunzip(out.file)
},.progress = "time")
```


# Data Pre-processing

Description: 

- Read in idat files and remove duplicated samples

Input: 

- idat files

Output: 

- RGSet.RDS
- phenoData
- BetaMatrixRaw

```{R, eval = FALSE}
RGSet <- read.metharray.exp(base = data.dir.raw,
                            recursive = TRUE,
                            verbose = TRUE) #dim: 622399 60
phenoData <- colData(se)
save(RGSet,phenoData, file = paste0(data.dir.read, "Gasparoni.rda")) 
```


# Data QC 

## Bilsufite conversion filtering

Removing samples with bisulfiteConversion lower than 88.

```{R qc_samples_packages, message = FALSE, results = "hide"}
library(wateRmelon)
library(RPMM)
#library(sesame)
#library(sesameData)
```


```{R}
load(file = paste0(data.dir.read, "Gasparoni.rda"))
bs <- data.frame(bisulfiteConversion = bscon(RGSet))
bsFilteredOut <- row.names(bs)[bs$bisulfiteConversion < 88]
nb.samples <- ncol(RGSet)
RGSet <- RGSet[,!colnames(RGSet) %in% bsFilteredOut]
nb.samples.bc.filtered <-  ncol(RGSet)
phenoData <- phenoData[match(substr(colnames(RGSet),1,10), phenoData$geo_accession),]
```

```{R, eval = FALSE}
save(RGSet,
     nb.samples,
     bs,
     phenoData,
     nb.samples.bc.filtered, 
     file = paste0(data.dir.bsfilter, "/RGSet_bsfiltered.rda"))
```

```{R, include = FALSE, eval = TRUE}
load(file = paste0(data.dir.bsfilter, "/RGSet_bsfiltered.rda"))
```

```{R, include = FALSE, eval=FALSE}
ggpubr::gghistogram(bs$bisulfiteConversion,xlab = "bisulfite Conversion" )
```

## Clinical data filtering

```{R}
dim(phenoData)

phenoData$braak_stage.ch1 <- phenoData$braak_stage.ch1 %>% as.numeric()
phenoData$age.ch1 <- phenoData$age.ch1 %>% as.numeric()
### Subset rows and columns
pheno_df <- phenoData  %>% as.data.frame() %>%
  dplyr::filter(
    source_name_ch1 == "Frontal Cortex" &
      !is.na(phenoData$braak_stage.ch1) &
      phenoData$characteristics_ch1 == "cell type: bulk"
  ) %>% dplyr::select(
    c(
      "geo_accession",
      "donor_id.ch1",
      "sentrix_id.ch1",
      "age.ch1",
      "Sex.ch1",
      "braak_stage.ch1"
    )
  )

### Rename vars
colnames(pheno_df) <- c(
  "sample", "subject.id", "slide", "age.brain", "sex", "stage"
)

dim(pheno_df) 
nb.samples.with.clinical <- nrow(pheno_df)
## phenotype dataset
save(RGSet,
     nb.samples.with.clinical,
     pheno_df,
     file = paste0(data.dir.clinical.filter, "/gasparoni_bs_and_clinical_filtered.rda"))
```

## Probes QC

Input: 

- RGSet.RDS
- beta_mat.RDS

Output: 

- beta_CG_XY_SNPfiltered.RDS


```{R}
##### 1. subset to probes with detection P <= 0.01 #############################
### Read in RGSet and beta_mat
load(paste0(data.dir.clinical.filter, "/gasparoni_bs_and_clinical_filtered.rda"))
```

```{R, message = FALSE, results = 'hide'}
### subset to probes with detection P <= 0.01
library(minfi)
```

```{R}
nb.probes <- nrow(RGSet)
detP <- detectionP(RGSet, type = "m+u")
failed.01 <- detP > 0.01
passedProbes <- rownames(failed.01)[rowMeans(failed.01) == 0] 

beta_mat <- getBeta(RGSet) 

beta_mat <- beta_mat[passedProbes, ]
nb.probes.detectP <- nrow(beta_mat)

##### 2. keep only probes that start with "cg" #################################
beta_mat <- beta_mat[grep("cg",rownames(beta_mat)),]
dim(beta_mat) 
nb.probes.detectP.cg <- nrow(beta_mat)

##### 3. drop probes that are on X/Y ###########################################
##### 4. drop probes where SNP with MAF >= 0.01 in the last 5 bp of the probe ##
```

```{R, message = FALSE, warning = FALSE}
library(DMRcate)
beta_mat <- rmSNPandCH(
  object = beta_mat,
  dist = 5, 
  mafcut = 0.01, 
  and = TRUE,
  rmcrosshyb = FALSE,
  rmXY = TRUE
) 
nb.probes.cg.dmrcate <- nrow(beta_mat)

##### 5. Output datasets #######################################################
save(
  pheno_df,
  nb.probes.detectP,
  nb.probes.detectP.cg,
  nb.probes.cg.dmrcate,
  beta_mat,
  file = paste0(data.dir.probes.qc, "beta_CG_XY_SNPfiltered_mat.rda")
)
```

# Normalization

- Quantile normalization and BMIQ normalization

Input: 

- beta_CG_XY_SNPfiltered_mat.RDS
- RGSet.RDS
- pheno_df.RDS
- full.annot.RDS

Output: 

- bs.csv
- pheno_df.RDS
- QNBMIQ.RDS

```{R qcSamplesData}
load(paste0(data.dir.probes.qc, "beta_CG_XY_SNPfiltered_mat.rda"))
```


## Quantile normalization

```{R, message = FALSE, results = 'hide'}
library(lumi)
```


```{R}
betaQN <- lumiN(x.lumi = beta_mat, method = "quantile")
dim(betaQN)
```

## BMIQ

```{R, message = FALSE, results = "hide"}
library(wateRmelon)
library(RPMM)
library(sesame)
library(sesameData)
```

```{R, eval = FALSE, include = FALSE}
### Order annotation in the same order as beta matrix
annotType <- sesameDataGet("HM450.hg19.manifest")
annotType$designTypeNumeric <- ifelse(annotType$designType == "I",1,2)

### Density plot for type I and type II probes
library(sm)

betaQNCompleteCol1 <- betaQN[complete.cases(betaQN[,1]), ]
annotTypeCompleteCol1 <- annotType[row.names(betaQNCompleteCol1), ]

sm.density.compare(
  betaQNCompleteCol1[,1],
  annotTypeCompleteCol1$designTypeNumeric
)

type12 <- annotType$designTypeNumeric[match(rownames(betaQN),names(annotType))]
```
```{R, eval = FALSE}
### BMIQ
set.seed (946)
doParallel::registerDoParallel(cores = 8)
betaQN_BMIQ <- plyr::aaply(
  betaQN, 2,
  function(x){
    norm_ls <- BMIQ(x, design.v = type12, plots = FALSE)
    return (norm_ls$nbeta)
  },.progress = "time",.parallel = TRUE
) %>% t()
colnames(betaQN_BMIQ) <- substr(colnames(betaQN_BMIQ),1,stringr::str_length(pheno_df$sample) %>% unique)

save(betaQN_BMIQ, pheno_df, file = paste0(data.dir.probes.normalization, "GASPARONI_QNBMIQ.rda"))
```

```{R, include = FALSE}
load(paste0(data.dir.probes.normalization, "GASPARONI_QNBMIQ.rda"))
```

```{R, eval = FALSE, include = FALSE}
##### 6. Density plot after normalization and BMIQ #############################
matdensity(object = betaQN_BMIQ,
           groupFactor =  pheno_df$stage %>%  as.factor(),
           xlab = " ", 
           ylab = "density",
           main = "Beta Values", 
           brewer.n = 8, 
           brewer.name = "Dark2")

legend('top',
       paste0("stage ", levels( pheno_df$stage %>%  as.factor())),
       col = c(1:7), lty = 1, lwd = 3)
```

# Outliers detection - PCA analysis

- Select most variable probes and perform PCA analysis

Input: 

- GASPARONI_QNBMIQ.rds
- pheno_df.RDS

Output: 

- GASPARONI_PCs_usingBetas.csv, 
- PCA plots
- GASPARONI_QNBMIQ_PCfiltered.RDS
- pheno_df.RDS


```{R}
# plotPCA and OrderDataBySd functions
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")
```

```{R}
### merge in group info
## add center and scale
## compare M values vs. beta values
## with and without center / scale

##### 0.Import datasets ########################################################
load(paste0(data.dir.probes.normalization, "GASPARONI_QNBMIQ.rda"))
identical(colnames(betaQN_BMIQ), pheno_df$sample)

### transform to m values
mvalue_mat <- log2(betaQN_BMIQ / (1 - betaQN_BMIQ)) #dim: 437713 142

pheno_df <- subset(pheno_df, pheno_df$sample %in% colnames(betaQN_BMIQ)) #dim: 142 7

# both_df$stage <- as.numeric(as.character(both_df$stage))

pheno_df$stage3 <- pheno_df$stage
pheno_df$stage3[pheno_df$stage <= 2] <- '0-2'
pheno_df$stage3[pheno_df$stage > 2 & pheno_df$stage < 5] <- '3-4'
pheno_df$stage3[pheno_df$stage >= 5] <- '5-6'


##### 1.Order matrix by most variable probes on top ############################

betaOrd_mat <- OrderDataBySd(betaQN_BMIQ) #dim: 391482 142

mOrd_mat <- OrderDataBySd(mvalue_mat)  #dim: 391482 142

betaOrd_matPlot <- betaOrd_mat[, pheno_df$sample] #dim: 391482 142
mOrd_matPlot <- mOrd_mat[, pheno_df$sample]       #dim: 391482 142
identical(pheno_df$sample, colnames(betaOrd_matPlot))
identical(pheno_df$sample, colnames(mOrd_matPlot))

expSorted_mat = betaOrd_mat #dim: 391482 142

pca <- prcomp(t(expSorted_mat[1:50000, ]),
              center = TRUE,
              scale = TRUE)

d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2])

meanPC1 <- mean (d$PC1)
sdPC1   <- sd (d$PC1)

meanPC2 <- mean (d$PC2)
sdPC2   <- sd (d$PC2)

out3sdPC1_1 <- meanPC1 - 3 * sdPC1
out3sdPC1_2 <- meanPC1 + 3 * sdPC1

out3sdPC2_1 <- meanPC2 - 3 * sdPC2
out3sdPC2_2 <- meanPC2 + 3 * sdPC2

d$outlier_PC1[d$PC1 >= out3sdPC1_1 & d$PC1 <= out3sdPC1_2] <- 0
d$outlier_PC1[d$PC1 < out3sdPC1_1 | d$PC1 > out3sdPC1_2] <- 1

d$outlier_PC2[d$PC2 >= out3sdPC2_1 & d$PC2 <= out3sdPC2_2] <- 0
d$outlier_PC2[d$PC2 < out3sdPC2_1 | d$PC2 > out3sdPC2_2] <- 1

readr::write_csv(d, paste0(data.dir.pca, "GASPARONI_PCs_usingBetas.csv"))
```

```{R, eval = FALSE, include = FALSE}
##### 2.PCA plot ###############################################################
library(ggplot2)
library(ggrepel)

### beta values
byStage <- plotPCA(
  dataset = "Gasparoni: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "stage3",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "Gasparoni: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)


#### M values
byStage <- plotPCA(
  dataset = "Gasparoni: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "stage3",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "Gasparoni: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
```


```{R}
##### 2.Filter samples by PCA, save files ######################################

noOutliers <- d[which(d$outlier_PC1 == 0 & d$outlier_PC2 == 0), ]
betaQN_BMIQ_PCfiltered <- betaQN_BMIQ[, rownames(noOutliers)] #dim: 433656 59
saveRDS(betaQN_BMIQ_PCfiltered, paste0(data.dir.pca, "Gasparoni_QNBMIQ_PCfiltered.RDS"))

pheno_df <- pheno_df[pheno_df$sample %in% rownames(noOutliers),] #dim: 59 6
saveRDS(pheno_df, paste0(data.dir.pca, "pheno_df.RDS"))
```

# Summary after QC steps

## Data and metadata
```{R}
nb.samples.with.clinical.after.pca <- ncol(betaQN_BMIQ_PCfiltered)
dim(betaQN_BMIQ_PCfiltered)
dim(pheno_df)
pheno_df %>% 
  DT::datatable(filter = 'top',
                style = "bootstrap",
                extensions = 'Buttons',
                options = list(scrollX = TRUE, 
                               dom = 'Bfrtip',
                               buttons = I('colvis'),
                               keys = TRUE, 
                               pageLength = 10), 
                rownames = FALSE,
                caption = "Samples metadata")
```

```{R, eval = FALSE, include = FALSE}
ggpubr::gghistogram(data = pheno_df, x = "stage",bins = 8)
ggpubr::gghistogram(data = pheno_df, x = "stage",bins = 8,facet.by = "sex",fill = "sex")
ggpubr::gghistogram(data = pheno_df, x = "age.brain",bins = 20)
ggpubr::gghistogram(data = pheno_df, x = "age.brain",bins = 20,fill = "sex",facet.by = "sex")
```


## Numbers of samples and probes removed in each step

```{R}
df.samples <- data.frame("Number of samples" =  c(nb.samples, 
                                                  nb.samples.bc.filtered,
                                                  nb.samples.with.clinical, 
                                                  nb.samples.with.clinical.after.pca),
                         "Description" = c("total number of samples",
                                           "samples with bisulfate conversion > 88",
                                           "samples with clinical data",
                                           "Samples after PCA"),
                         "Difference" = c("-",
                                          nb.samples.bc.filtered - nb.samples ,
                                          nb.samples.with.clinical - nb.samples.bc.filtered ,
                                          nb.samples.with.clinical.after.pca - nb.samples.with.clinical)
)    
df.samples           

# Create summary table
df.probes <- data.frame("Number of probes" = c(nb.probes,
                                               nb.probes.detectP, 
                                               nb.probes.detectP.cg,
                                               nb.probes.cg.dmrcate),
                        "Description" = c("total number of probes in raw data",
                                          "detection P < 0.01",
                                          "only probes that start with cg",
                                          "DMRcate"),
                        "Difference" = c("-",
                                         nb.probes.detectP - nb.probes ,
                                         nb.probes.detectP.cg - nb.probes.detectP,
                                         nb.probes.cg.dmrcate - nb.probes.detectP.cg)
)
df.probes

save(df.samples,df.probes,file = file.path(data.dir.table, "GASPARONI_table.rda"))
```

# Compute neuron proportion

Data from  https://www.tandfonline.com/doi/full/10.4161/epi.23924

- Input: Gasparoni_QNBMIQ_PCfiltered.RDS, pheno_df.RDS

- Output: pheno_withNeuronProp_df.RDS

```{R}
objects <- load("../../CET/CETS_Image.RData")
objects
```

## Get reference profile from Caucasions + controls 
```{R}
idx <- list(
  controlNeuron = pdBrain$celltype == "N" & pdBrain$diag == "Control" & pdBrain$ethnicity == "Caucasian",
  controlGlia   = pdBrain$celltype == "G" & pdBrain$diag == "Control" & pdBrain$ethnicity == "Caucasian"
)

refProfile <- getReference(brain, idx)


##### 2. Estimate proportions of neurons in PFC samples ########################

### Limit to 10,000 cpgs in the refProfile dataset
pfc <- readRDS(paste0(data.dir.pca, "Gasparoni_QNBMIQ_PCfiltered.RDS")) #dim: 433656 59

selected <- rownames(pfc) %in% rownames(refProfile)

pfc.refcpgs <- pfc[selected, ] 

### Estimate proportion of neurons
prop <- data.frame(estProportion(pfc.refcpgs, profile = refProfile))
colnames(prop) <- "prop.neuron"

##### 3. Merge pfc.refcpgs with phenotype file #################################
pheno <- readRDS(paste0(data.dir.pca, "pheno_df.RDS"))

pheno_final <- merge(
  pheno,
  prop,
  by.x = "sample",
  by.y = "row.names"
)

saveRDS(pheno_final, paste0(data.dir.neuron, "pheno_withNeuronProp_df.RDS"))
```

# Linear regression by cpgs Methylation 

Input: 

- Gasparoni_QNBMIQ_PCfiltered.RDS,
- pheno59_withNeuronProp_df.RDS

Output:

- Gasparoni_single_cpg_pVal_df.csv

## Import datasets

```{R}
beta_mat <- readRDS(paste0(data.dir.pca, "Gasparoni_QNBMIQ_PCfiltered.RDS")) 
pheno_df <- readRDS(paste0(data.dir.neuron, "pheno_withNeuronProp_df.RDS")) 
```

## Test all cpgs

```{R, eval = TRUE}
### Compute M values
mval_mat <- log2(beta_mat / (1 - beta_mat))

pheno_df$Sample <- pheno_df$sample

identical(pheno_df$Sample, colnames(mval_mat))

pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)
# If rosmap cohort, don't forget batch effect

str(pheno_df)

is(pheno_df$stage,"numeric")
is(pheno_df$age.brain,"numeric")
is(pheno_df$prop.neuron,"numeric")
```


```{R, eval = FALSE}
predictors_char <- "stage"
covariates_char <- c("age.brain", "sex", "prop.neuron", "slide")


doParallel::registerDoParallel(cores = parallel::detectCores()/2)
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

results_ordered_df <- plyr::adply(mval_mat,1, function(row){
  
  sumOneRegion_df <- data.frame(t(row))
  
  result <- TestSingleRegion(
    predictors_char = predictors_char,
    covariates_char = covariates_char,
    pheno_df = pheno_df,
    sumOneRegion_df = sumOneRegion_df
  )
  result
}, .progress = "time",.parallel = TRUE,.id = "cpg")
colnames(results_ordered_df)[1] <- "cpg"

identical(row.names(mval_mat), results_ordered_df$cpg %>% as.character())

results_ordered_df$fdr <- p.adjust(
    results_ordered_df$pValue,
    method = "fdr"
)

write.csv(
  results_ordered_df,
  paste0(data.dir.single.cpg.pval, "Gasparoni_single_cpg_pVal_df.csv"),
  row.names = FALSE
)
```

```{R}
results_ordered_df <- readr::read_csv(
  paste0(data.dir.single.cpg.pval, "Gasparoni_single_cpg_pVal_df.csv"),
  col_types = readr::cols())
results_ordered_df
```


# Linear regression by regions median Methylation 

## Residuals control and coMethylated Regions

1. Take residuals
2. Find co-methylated regions

Input: 

- QNBMIQ_PCfiltered
- pheno_withNeuronProp_df

Output: 

- QNBMIQ_PCfiltered_mvalResiduals
- residuals_cometh_ls

### Residuals

```{R, eval = FALSE}
##### 1. Import datasets #######################################################
beta_mat <- readRDS(paste0(data.dir.pca, "Gasparoni_QNBMIQ_PCfiltered.RDS")) #dim:433656 59
pheno_df <- readRDS(paste0(data.dir.neuron, "pheno_withNeuronProp_df.RDS")) #dim:59 7

### Compute M values
mvalue_mat <- log2(beta_mat / (1 - beta_mat))

### Reorder samples based on pheno_df
mvalue_mat <- mvalue_mat[, pheno_df$sample]

identical(colnames(mvalue_mat),  pheno_df$sample)

### Take residuals
lmF <- function(mval){
  fitE <- lm(
    as.numeric(mval) ~ age.brain + sex + prop.neuron + as.character(slide), #add batch if rosmap
    data = pheno_df,
    na.action = na.exclude
  )
  residuals (fitE)
}
doParallel::registerDoParallel(cores = 4)
resid <- plyr::adply(mvalue_mat,1,.fun = lmF,.progress = "time",.parallel = TRUE)
rownames(resid) <- resid[,1]
resid[,1] <- NULL
colnames(resid) <- colnames(mvalue_mat)
dim(resid)
dim(mvalue_mat)

### Save dataset
saveRDS(
  resid,
  paste0(data.dir.residuals, "Gasparoni_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)
```

### Find co-methylated regions

```{R, eval = FALSE}
### Import datasets
mvalue_residuals_mat <- readRDS(
  paste0(data.dir.residuals, "Gasparoni_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)

### Call in functions
library(coMethDMR)
library(BiocParallel)

probes.cluster.all <- coMethDMR::getPredefinedCluster(arrayType = "450k",
                                                      clusterType = "regions")

ncores <- parallel::detectCores()/2
### Find co-methylated clusters
coMeth_ls <- CoMethAllRegions(
  dnam = mvalue_residuals_mat,      
  betaToM = FALSE,                   
  CpGs_ls = probes.cluster.all,
  arrayType = "450k",
  rDropThresh_num = 0.4,
  minPairwiseCorr = NULL,
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "all",
  nCores_int = ncores,
  progressbar = FALSE
)

saveRDS(
  coMeth_ls,
  paste0(data.dir.residuals,"Gasparoni_residuals_cometh_input_ls.RDS")
)
```

## Linear regression by regions median Methylation 

1. Calculate medians by cluster and sample
2. linear regression

Input: 

- QNBMIQ_PCfiltered,
- pheno_withNeuronProp_df
- residuals_cometh_input_ls

Output: 

- info_df
- mediansMval_df
- linear_df_aging

### Calculate medians by cluster and sample

```{R, eval = FALSE}
### Import datasets
beta_mat <- readRDS(paste0(data.dir.pca,
                           "Gasparoni_QNBMIQ_PCfiltered.RDS"))
pheno_df <- readRDS(paste0(data.dir.neuron, "pheno_withNeuronProp_df.RDS"))
mval_mat <- log2(beta_mat / (1 - beta_mat)) %>% as.matrix()
coMeth_ls <- readRDS(
  paste0(data.dir.residuals, "Gasparoni_residuals_cometh_input_ls.RDS")
)

### Create info dataset
input_cometh <- data.frame(
  inputRegion = coMeth_ls$inputRegion_chr,
  nCoMethRegion = coMeth_ls$nCoMethRegions_num,
  coMethRegion = names(coMeth_ls$coMeth_ls),
  nCpGs = unlist(lapply(coMeth_ls$coMeth_ls, length), use.names = FALSE),
  stringsAsFactors = FALSE
)

input_cometh_nodup <- input_cometh[
  !duplicated(input_cometh$coMethRegion),
  ]
colnames(input_cometh_nodup) <- c(
  paste0(cohort, "_inputRegion"),
  paste0(cohort, "_nCoMethRegion"),
  paste0(cohort, "_coMethRegion"),
  paste0(cohort, "_nCpGs")
)

saveRDS(
  input_cometh_nodup,
  paste0(data.dir.median, cohort,"_info_df.rds")
)

### Take median of probes in each cluster for each sample
filename <-  paste0(paste0(data.dir.median, cohort, "_mediansMval_df.rds"))
library(robustbase)
mval_mat <- mval_mat[rownames(mval_mat) %in% unlist(coMeth_ls$coMeth_ls),]
if(!file.exists(filename)){
  medianMval.df <- plyr::ldply(
    coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))],
    function(probes){
      colMedians(mval_mat[as.character(probes),], na.rm = TRUE)
    },
    .progress = "time"
  )
  medianMval.df$.id <- NULL
  colnames(medianMval.df) <- colnames(mval_mat)
  saveRDS(medianMval.df, file = filename)
} else {
  medianMval.df <- readRDS(filename)
}
```

### Test all regions -- linear regressions

```{R, eval = TRUE}
### Import datasets
cohort <- "Gasparoni"
info_df <- readRDS(dir(data.dir.median, pattern = "info", full.names = TRUE))
mediansMval_df <- readRDS(dir(data.dir.median, pattern = "mediansMval", full.names = TRUE))
pheno_df <- readRDS(paste0(data.dir.neuron, "pheno_withNeuronProp_df.RDS")) 

### Check variables before fitting model
pheno_df$Sample <- pheno_df$sample
identical(pheno_df$Sample, colnames(mediansMval_df))

pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)
# If rosmap cohort, don't forget batch effect

str(pheno_df)
```

```{R, eval = FALSE}
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

predictors_char <- "stage"
covariates_char <- c("age.brain", "sex", "prop.neuron", "slide")

res_df <- TestAllRegions_noInfo(
  predictors_char = predictors_char,
  covariates_char = covariates_char,
  pheno_df = pheno_df,
  summarizedRegions_df = mediansMval_df
)

colnames(res_df) <- c(
  paste0(cohort, "_estimate"),
  paste0(cohort, "_se"),
  paste0(cohort, "_pVal"),
  paste0(cohort, "_fdr")
)

res_withInfo_df <- cbind(info_df, res_df)

saveRDS(
  res_withInfo_df,
  paste0(data.dir.median,  cohort, "_linear_df.rds")
)
```

```{R}
file <- dir(data.dir.median,pattern = paste0(".*linear_df"),
            recursive = T,
            full.names = TRUE,
            ignore.case = T)
file
res_withInfo_df <- readRDS(file)
dim(res_withInfo_df)
res_withInfo_df
```

# Data final

```{R}
dir(path = data.dir,recursive = T, pattern = ".rda|.csv|.RDS")
```

# Session information
```{R}
devtools::session_info()
```
