---
title: "LONDON blood dataset"
author: "Lanyu Zhang, Tiago C. Silva, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: lumen
    highlight: zenburn
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```


# Data retrival

```{R, message = FALSE, results = 'hide'}
library(dplyr)
```

```{R}
cohort <- "LONDON_blood"
data.dir <- file.path("DATASETS/",cohort,"/") 
data.dir.table <- "DATASETS/Summary_Table/" 
data.dir.raw <- file.path(data.dir,"/step1_download/") 
data.dir.clinical.filter <- file.path(data.dir,"/step2_clinical_available_filtering/") 
data.dir.probes.qc <- file.path(data.dir,"/step3_probesQC_filtering/") 
data.dir.probes.normalization <- file.path(data.dir,"/step4_normalization/") 
data.dir.pca <- file.path(data.dir,"/step5_pca_filtering/") 
data.dir.neuron <- file.path(data.dir,"/step6_neuron_comp/") 
# data.dir.single.cpg.pval <- file.path(data.dir,"/step7_single_cpg_pval/") 
data.dir.residuals <- file.path(data.dir,"/step7_residuals/") 
data.dir.median <- file.path(data.dir,"/step8_median/")
data.dir.validation <- file.path(data.dir,"/step9_validation/") 
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

Required R library `GEOquery` can be installed as following:

```{R, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("GEOquery")
```

Source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59685

```{r synapse, eval = FALSE}
library(GEOquery)
library(SummarizedExperiment)
test <- getGEO(GEO = "GSE59685",destdir = data.dir,GSEMatrix = TRUE,)
GSE59685 <- test$GSE59685_series_matrix.txt.gz %>% makeSummarizedExperimentFromExpressionSet() 
metadata <- colData(GSE59685)
## Failed to create assayData with pacakage GEOquery, so we use the beta matrix
## directly downloaded online instead
library(data.table)
assayData <- fread(
  "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE59nnn/GSE59685/suppl/GSE59685_betas.csv.gz",
  skip = 5,
  header = TRUE,
  sep =','
)

## Turn assayData from "data.table" and "data.frame" to "data.frame" only
assayData <- data.frame(assayData)

## Exclude row 1 as it's old sample id
assayData <- assayData[-1, ]

## Create row names based on V1
row.names(assayData) <- assayData$V1

## Delete column V1 after creating rownames
assayData <- assayData[, 2:ncol(assayData)]

assayData <- data.matrix(assayData)

## Save RDS file
saveRDS(assayData, paste0(data.dir.raw, "GSE59685_assay.RDS"))

### Extract and save phenotype dataset
phenoData <-  colData(GSE59685)
phenoData <- phenoData[match(colnames(assayData),rownames(phenoData)),]
saveRDS(phenoData, paste0(data.dir.raw, "GSE59685_pheno.RDS"))
write.csv(phenoData, paste0(data.dir.raw, "GSE59685_pheno.csv"))
```

#  Data Pre-processing 

Description:

 Subset Files
 1. subset pheno data to
  1) whole blood tissue and status != "Exclude" (n = 57)
  2) variables -- sample, subject_id, sentrix_id, slide, age, sex, stage
 2. subset methylation data to match subjects in pheno data (n = 57)

 Input: GSE59685_pheno.csv, GSE59685_assay.RDS
 Output: pheno_BLOOD_df.RDS, beta110_BLOOD_mat.RDS



```{R, eval = TRUE}
##### 1. Subset pheno data #####################################################

### Read in phenotype data
phenoRaw_df <- readr::read_csv(paste0(data.dir.raw, "GSE59685_pheno.csv")) #dim: 531 49

### Subset rows and columns

### Subset rows and columns
phenoBLOOD_df <- phenoRaw_df[
  (phenoRaw_df$source_name_ch1 == "whole blood") &
    (phenoRaw_df$ad.disease.status.ch1 != "Exclude"),
  c("geo_accession", "subjectid.ch1", "barcode.ch1",
    "age.blood.ch1", "Sex.ch1", "ad.disease.status.ch1", "braak.stage.ch1")
]

### Rename vars
colnames(phenoBLOOD_df) <- c(
  "sample", "subject.id", "sentrix_id", "age.blood", "sex", "status", "stage"
)

### Turn factors into characters
phenoBLOOD_df$sample <- as.character(phenoBLOOD_df$sample)
phenoBLOOD_df$subject.id <- as.character(phenoBLOOD_df$subject.id)
phenoBLOOD_df$sentrix_id <- as.character(phenoBLOOD_df$sentrix_id)
phenoBLOOD_df$sex <- as.character(phenoBLOOD_df$sex)
phenoBLOOD_df$status <- as.character(phenoBLOOD_df$status)
phenoBLOOD_df$stage <- as.integer(as.character(phenoBLOOD_df$stage))

### Get slide from sentrix_id
  # e.g. "6042316048_R05C01"(sentrix_id) -- "6042316048"(slide) and "R05C01"(array)
sentrixID_mat <- do.call(rbind, strsplit(phenoBLOOD_df$sentrix_id, "_"))
phenoBLOOD_df$slide <- sentrixID_mat[, 1]

### Order final pheno_df
pheno_df <- phenoBLOOD_df[, c(
  "sample", "subject.id", "sentrix_id", "slide",
  "age.blood", "sex", "status", "stage"
)] 
```

```{R}
##### 2. Subset methylation data ###############################################

### Read in methylation data
beta_mat <- readRDS(paste0(data.dir.raw, "GSE59685_assay.RDS")) #dim: 485577 531

### Subset methylation data based on pheno data
beta_mat <- beta_mat[, match(pheno_df$sample,colnames(beta_mat))] # dim: 485577 110

##### 3. Output datasets #######################################################

## phenotype dataset
saveRDS(pheno_df, paste0(data.dir.clinical.filter, "pheno_BLOOD_df.RDS")) 

## methylation beta values dataset
saveRDS(beta_mat, paste0(data.dir.clinical.filter, "beta_BLOOD_mat.RDS")) 
```

## Probes QC

 1. keep only probes that start with "cg"
 2. drop probes that are on X/Y
 3. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.

 Input: beta110_PFC_mat.RDS
 
 Output: beta110_PFC_CG_XY_SNPfiltered_mat.RDS

```{R}
##### 1. keep on probes with start with "cg" ###################################

beta_mat <- readRDS(paste0(data.dir.clinical.filter, "beta_BLOOD_mat.RDS")) #dim: 485577 110
nb.probes <- nrow(beta_mat)
nb.samples <- ncol(beta_mat)
nb.samples.with.clinical <- ncol(beta_mat)

beta_mat <- beta_mat[grep("cg",rownames(beta_mat)),]
dim(beta_mat)
nb.probes.cg <- nrow(beta_mat)
```

```{R, message = FALSE, results = "hide"}
##### 2. drop probes that are on X/Y ###########################################
##### 3. drop probes where SNP with MAF >= 0.01 in the last 5 bp of the probe ##
library(DMRcate)
```

```{R, message = FALSE, warning = FALSE}
beta_CG_XY_SNPfiltered_mat <- rmSNPandCH(
  object = beta_mat,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = FALSE,
  rmXY = TRUE
)
dim(beta_CG_XY_SNPfiltered_mat)

##### 4. Output datasets #######################################################

saveRDS(
  beta_CG_XY_SNPfiltered_mat,
  paste0(data.dir.probes.qc, "beta_CG_XY_SNPfiltered_mat.RDS")
)

```

## Samples QC

 1. quantro test - decide if quantile normalization is needed
 2. quantile normalization - box plots before and after sample QC
 3. BMIQ - density plots before and after sampel QC

 Input: 
 
 - beta_CG_XY_SNPfiltered_mat.RDS
 - pheno_PFC_df.RDS
 - full.annot.RDS
 
 Output: 
 
- London_PFC_QNBMIQ.RDS,
- boxplots before and after quantile normalization
- denstity plots before and after quantile normalization and BMIQ


### Quantro test

```{R}
beta_mat <- readRDS(paste0(data.dir.probes.qc, "beta_CG_XY_SNPfiltered_mat.RDS"))
pheno_df <- readRDS(paste0(data.dir.clinical.filter, "pheno_BLOOD_df.RDS")) 
nb.probes.cg.dmrcate <- nrow(beta_mat)
```


```{R, results = "hide", message = FALSE}
library(quantro)
library(doParallel)
```

```{R, eval = FALSE}
cl <- makeCluster(4)
registerDoParallel(cl)
qtestPerm <- quantro(object = beta_mat[complete.cases(beta_mat),],
                     groupFactor = pheno_df$status,
                     B = 1000)
stopCluster(cl)
save(qtestPerm,file = file.path(data.dir.probes.normalization,"qtestPerm.rda"))
```

```{R, include = FALSE, eval = TRUE}
load(file.path(data.dir.probes.normalization,"qtestPerm.rda"))
```

```{R}
qtestPerm
```

### Plot before normalization and BMIQ

```{R}
### Density plot
matdensity(object = beta_mat,
           groupFactor = pheno_df$status %>% as.factor,
           xlab = " ", ylab = "density",
           main = "Beta Values", brewer.n = 8, brewer.name = "Dark2")

legend('top',
       paste0("status ", levels(pheno_df$status %>% as.factor)),
       col = c(1:7), lty = 1, lwd = 3)

matboxplot(beta_mat,
           groupFactor = pheno_df$status,
           xaxt = "n",
           main = "Beta Values in London Blood cohort (autosomes) - before normalization")

legend('bottom',
       paste0("status ",  levels(pheno_df$status %>% as.factor)),
       col = c(1:7), lty = 1, lwd = 3, cex = 0.70)
```

### Quantile normalization

```{R}
library(lumi)
betaQN <- lumiN(x.lumi = beta_mat, method = "quantile")

dim(betaQN)
```


### Boxplot after normalization
```{R}
# pheno_df$stage <- factor(pheno_df$stage)
matboxplot(betaQN,
           groupFactor = pheno_df$status,
           xaxt = "n",
           main = "Beta Values in London Blood cohort (autosomes) - after normalization")

legend('bottom',
       paste0("status ", levels(pheno_df$status %>% as.factor)),
       col = c(1:7), lty = 1, lwd = 3, cex = 0.70)
```

```{R, message = FALSE, results = "hide"}
##### 5. BMIQ ##################################################################
library(wateRmelon)
library(RPMM)
library(sesame)
library(sesameData)
```

```{R}
### Order annotation in the same order as beta matrix
annotType <- sesameDataGet("HM450.hg19.manifest")
annotType$designTypeNumeric <- ifelse(annotType$designType == "I",1,2)

### Density plot for type I and type II probes
library(sm)

betaQNCompleteCol1 <- betaQN[complete.cases(betaQN[,1]), ]
annotTypeCompleteCol1 <- annotType[row.names(betaQNCompleteCol1), ]

sm.density.compare(
  betaQNCompleteCol1[,1],
  annotTypeCompleteCol1$designTypeNumeric
)

type12 <- annotType$designTypeNumeric[match(rownames(betaQN),names(annotType))]
```

```{R, eval = FALSE}
### BMIQ
set.seed (946)
doParallel::registerDoParallel(cores = 8)
betaQN_BMIQ <- plyr::aaply(
    betaQN, 2,
    function(x){
        norm_ls <- BMIQ(x, design.v = type12, plots = FALSE)
        return (norm_ls$nbeta)
    },.progress = "time",.parallel = TRUE
) %>% t()

saveRDS(betaQN_BMIQ, paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ.RDS"))
```


```{R, include = FALSE}
betaQN_BMIQ <- readRDS(paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ.RDS"))
```


```{R}
##### 6. Density plot after normalization and BMIQ #############################
# pheno_df$stage <- factor(pheno_df$stage)
matdensity(object = betaQN_BMIQ,
           groupFactor = pheno_df$status,
           xlab = " ", ylab = "density",
           main = "Beta Values", brewer.n = 8, brewer.name = "Dark2")

legend('top',
       paste0("status ", levels(pheno_df$status %>% as.factor)),
       col = c(1:7), lty = 1, lwd = 3)

```


# Outliers detection - PCA analysis


Description: 

1. estimate standard deviation for each probe
2. select most variable probes (e.g. n = 50,000)
3. pca plot
4. Filter outliers

Input: 

- QNBMIQ.rds
- pheno_df.RDS

Output: 

- PCs_usingBetas.csv, 
- PCA plots
- QNBMIQ_PCfiltered.RDS
- pheno_df.RDS


```{R}
# plotPCA and OrderDataBySd functions
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

beta_mat <- readRDS(paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ.RDS")) #dim: 437713 110

pheno_df <- readRDS(paste0(data.dir.clinical.filter, "pheno_BLOOD_df.RDS")) #dim: 110 7

identical(colnames(beta_mat), pheno_df$sample)

### transform to m values
mvalue_mat <- log2(beta_mat/(1 - beta_mat)) #dim: 437713 110

pheno_df <- subset(pheno_df, pheno_df$sample %in% colnames(beta_mat)) #dim: 110 7
##### 1.Order matrix by most variable probes on top ############################

betaOrd_mat <- OrderDataBySd(beta_mat) #dim: 437713 110

mOrd_mat <- OrderDataBySd(mvalue_mat)  #dim: 437713 110

betaOrd_matPlot <- betaOrd_mat[, pheno_df$sample] #dim: 437713 110
mOrd_matPlot <- mOrd_mat[, pheno_df$sample]       #dim: 437713 110
identical(pheno_df$sample, colnames(betaOrd_matPlot))
identical(pheno_df$sample, colnames(mOrd_matPlot))

expSorted_mat = betaOrd_mat #dim: 437713 110

pca <- prcomp(
  t(expSorted_mat[1:50000,]),
  center = TRUE,
  scale = TRUE
)


d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2])

meanPC1 <- mean (d$PC1)
sdPC1   <- sd (d$PC1)

meanPC2 <- mean (d$PC2)
sdPC2   <- sd (d$PC2)

out3sdPC1_1 <- meanPC1 - 3*sdPC1
out3sdPC1_2 <- meanPC1 + 3*sdPC1

out3sdPC2_1 <- meanPC2 - 3*sdPC2
out3sdPC2_2 <- meanPC2 + 3*sdPC2


d$outlier_PC1[d$PC1 >= out3sdPC1_1 & d$PC1 <= out3sdPC1_2] <- 0
d$outlier_PC1[d$PC1 < out3sdPC1_1 | d$PC1 > out3sdPC1_2] <- 1

d$outlier_PC2[d$PC2 >= out3sdPC2_1 & d$PC2 <= out3sdPC2_2] <- 0
d$outlier_PC2[d$PC2 < out3sdPC2_1 | d$PC2 > out3sdPC2_2] <- 1

write.csv(d, paste0(data.dir.pca, "London_Blood_PCs_usingBetas.csv"))


##### 2.PCA plot ###############################################################
library(ggplot2)
library(ggrepel)

### beta values
byStatus <- plotPCA(
  dataset = "London Blood: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "status",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "London Blood: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

### M values
byStatus <- plotPCA(
  dataset = "London Blood: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "status",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "London Blood: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
```

## Filter samples by PCA 
```{R, eval = FALSE}
noOutliers <- d[which(d$outlier_PC1 == 0 & d$outlier_PC2 == 0), ]
betaQN_BMIQ_PCfiltered <- beta_mat[, rownames(noOutliers)]
saveRDS(betaQN_BMIQ_PCfiltered, paste0(data.dir.pca, "London_QNBMIQ_PCfiltered.RDS"))

pheno_df <- pheno_df[pheno_df$sample %in% rownames(noOutliers),] 
saveRDS(pheno_df, paste0(data.dir.pca, "pheno_df.RDS"))
```

# Summary after QC steps

## Data and metadata
```{R}
betaQN_BMIQ_PCfiltered <- readRDS(paste0(data.dir.pca, "London_QNBMIQ_PCfiltered.RDS")) 
nb.samples.with.clinical.after.pca <- ncol(betaQN_BMIQ_PCfiltered)
pheno_df <- readRDS(paste0(data.dir.pca, "pheno_df.RDS"))
dim(betaQN_BMIQ_PCfiltered)
dim(pheno_df)
```

```{R}
pheno_df %>% 
  DT::datatable(filter = 'top',
                style = "bootstrap",
                extensions = 'Buttons',
                options = list(scrollX = TRUE, 
                               dom = 'Bfrtip',
                               buttons = I('colvis'),
                               keys = TRUE, 
                               pageLength = 10), 
                rownames = FALSE,
                caption = "Samples metadata")
```


```{R}
ggpubr::gghistogram(data = pheno_df, 
                    x = "status",
                    bins = 8, 
                    stat = "count")
ggpubr::gghistogram(data = pheno_df,
                    x = "status",
                    bins = 8,
                    facet.by = "sex",
                    fill = "sex",
                    stat = "count")
```



## Numbers of samples and probes removed in each step


```{R}
df.samples <- data.frame("Number of samples" =  c(nb.samples, 
                                                  nb.samples.with.clinical, 
                                                  nb.samples.with.clinical.after.pca),
                         "Description" = c("total number of samples",
                                           "samples with clinical data",
                                           "Samples after PCA"),
                         "Difference" = c("-",
                                          nb.samples.with.clinical - nb.samples ,
                                          nb.samples.with.clinical.after.pca - nb.samples.with.clinical)
)    
df.samples                     
# Create summary table
df.probes <- data.frame("Number of probes" = c(nb.probes,
                                               nb.probes.cg, 
                                               nb.probes.cg.dmrcate),
                        "Description" = c("total number of probes in raw data",
                                          "only probes that start with cg",
                                          "DMRcate"),
                        "Difference" = c("-",
                                         nb.probes.cg - nb.probes ,
                                         nb.probes.cg.dmrcate - nb.probes.cg)
)
df.probes
save(df.samples,df.probes,file = file.path(data.dir.table, "LONDON_blood_table.rda"))
```

# Compute neuron proportion


Data from  https://www.tandfonline.com/doi/full/10.4161/epi.23924

- Input: London_QNBMIQ_PCfiltered.RDS, pheno_blood_df.RDS
- Output: pheno_BLOOD_withBloodProp_df.rds

```{R}
blood <- readRDS(paste0(data.dir.pca, "London_QNBMIQ_PCfiltered.RDS")) 
nb.samples.with.clinical.after.pca <- ncol(blood)
pheno <- readRDS(paste0(data.dir.pca, "pheno_df.RDS"))
library(EpiDISH)

data(centDHSbloodDMC.m)

out.l <- epidish(blood, centDHSbloodDMC.m, method = 'RPC')
frac.m <- data.frame(out.l$estF)

pheno_final <- merge(
  pheno, 
  frac.m,
  by.x = "sample",
  by.y = "row.names",
  sort = FALSE
)
identical(pheno_final$sample, colnames(blood))

saveRDS(
  pheno_final,
  paste0(data.dir.neuron, "pheno_BLOOD_withBloodProp_df.rds")
)
```

# Linear regression by regions median Methylation 

## Residuals control and coMethylated Regions


1. Take residuals
2. Find co-methylated regions

Input: 

- QNBMIQ_PCfiltered
- pheno_withNeuronProp_df

Output: 

- QNBMIQ_PCfiltered_mvalResiduals
- residuals_cometh_ls

### Take residuals

```{R, eval = TRUE}
##### 1. Import datasets #######################################################
beta_mat <- readRDS(grep("QNBMIQ",dir(data.dir.pca,full.names = T),ignore.case = T,value = T)) 
pheno_df <- readRDS(dir(data.dir.neuron,full.names = T)) 

### Compute M values
mvalue_mat <- log2(beta_mat / (1 - beta_mat))

### Reorder samples based on pheno_df
mvalue_mat <- mvalue_mat[, pheno_df$sample]

identical(colnames(mvalue_mat),  pheno_df$sample)
```

```{R, eval = FALSE}
lmF <- function(mval){
  fitE <- lm(
    as.numeric(mval) ~ age.blood + sex + slide +
      B + NK + CD4T + CD8T + Mono + Neutro + Eosino,
    data = pheno_df,
    na.action = na.exclude
  )
  residuals (fitE)
}

doParallel::registerDoParallel(cores = 16)
resid <- plyr::adply(mvalue_mat,1,.fun = lmF,.progress = "time",.parallel = TRUE)
rownames(resid) <- resid[,1]
resid[,1] <- NULL
colnames(resid) <- colnames(mvalue_mat)

saveRDS(
  resid,
  paste0(data.dir.residuals,"LONDON_blood_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)
```

```{R, eval = FALSE}
##### 4. Find co-methylated regions ############################################
### Import datasets
mvalue_residuals_mat <- readRDS(
  paste0(data.dir.residuals, "LONDON_blood_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)

### Call in functions
library(coMethDMR)
library(BiocParallel)

probes.cluster.all <- coMethDMR::getPredefinedCluster(arrayType = "450k",
                                                      clusterType = "regions")

ncores <- 10
### Find co-methylated clusters
coMeth_ls <- CoMethAllRegions(
  dnam = mvalue_residuals_mat,      
  betaToM = FALSE,                   
  CpGs_ls = probes.cluster.all,
  arrayType = "450k",
  rDropThresh_num = 0.4,
  minPairwiseCorr = NULL,
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "all",
  nCores_int = ncores,
  progressbar = FALSE
)

saveRDS(
  coMeth_ls,
  paste0(data.dir.residuals,"LONDON_blood_residuals_cometh_input_ls.RDS")
)
```


## Linear regression by regions median Methylation 

1. Calculate medians by cluster and sample
2. linear regression

Input: 

- QNBMIQ_PCfiltered,
- pheno_withNeuronProp_df
- residuals_cometh_input_ls

Output: 

- info_df
- mediansMval_df
- linear_df_aging

### Calculate medians by cluster and sample

```{R, eval = FALSE}
cohort <- "LONDON_blood"

### Import datasets
beta_mat <- readRDS(dir(data.dir.pca, pattern = "QNBMIQ", full.names = TRUE))
pheno_df <- readRDS(dir(data.dir.neuron,full.names = T)) 
mval_mat <- log2(beta_mat / (1 - beta_mat)) %>% as.matrix()
coMeth_ls <- readRDS(
  paste0(data.dir.residuals, "LONDON_blood_residuals_cometh_input_ls.RDS")
)

### Create info dataset
input_cometh <- data.frame(
  inputRegion = coMeth_ls$inputRegion_chr,
  nCoMethRegion = coMeth_ls$nCoMethRegions_num,
  coMethRegion = names(coMeth_ls$coMeth_ls),
  nCpGs = unlist(lapply(coMeth_ls$coMeth_ls, length), use.names = FALSE),
  stringsAsFactors = FALSE
)

input_cometh_nodup <- input_cometh[
  !duplicated(input_cometh$coMethRegion),
]
colnames(input_cometh_nodup) <- c(
  paste0(cohort, "_inputRegion"),
  paste0(cohort, "_nCoMethRegion"),
  paste0(cohort, "_coMethRegion"),
  paste0(cohort, "_nCpGs")
)

saveRDS(
  input_cometh_nodup,
  paste0(data.dir.median,  cohort, "_info_df.rds")
)

### Take median of probes in each cluster for each sample
filename <-  paste0(paste0(data.dir.median, cohort, "_mediansMval_df.rds"))

library(robustbase)
mval_mat <- mval_mat[rownames(mval_mat) %in% unlist(coMeth_ls$coMeth_ls),]
if(!file.exists(filename)){
  medianMval.df <- plyr::ldply(
    coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))],
    function(probes){
      colMedians(mval_mat[as.character(probes),], na.rm = TRUE)
    },
    .progress = "time"
  )
  medianMval.df$.id <- NULL
  colnames(medianMval.df) <- colnames(mval_mat)
  saveRDS(medianMval.df, file = filename)
} else {
  medianMval.df <- readRDS(filename)
}
```

### Test all regions -- linear regressions

```{R, eval = TRUE}
### Import datasets
cohort <- "LONDON_blood"
info_df <- readRDS(dir(data.dir.median,pattern = "info", full.names = TRUE))
mediansMval_df <- readRDS(dir(data.dir.median,pattern = "mediansMval", full.names = TRUE))
pheno_df <- readRDS(dir(data.dir.neuron, full.names = TRUE))

### Check variables before fitting model
pheno_df$Sample <- pheno_df$sample

identical(pheno_df$Sample, colnames(mediansMval_df))

pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)

str(pheno_df)
```

```{R, eval = FALSE}
# If rosmap cohort, don't forget batch effect
predictors_char <- "status"
covariates_char <- c(
  "age.blood", "sex", "slide",
  "B", "NK", "CD4T", "CD8T", "Mono", "Neutro", "Eosino"
)
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

res_df <- TestAllRegions_noInfo(
  predictors_char = predictors_char,
  covariates_char = covariates_char,
  pheno_df = pheno_df,
  summarizedRegions_df = mediansMval_df
)

colnames(res_df) <- c(
  paste0(cohort, "_estimate"),
  paste0(cohort, "_se"),
  paste0(cohort, "_pVal"),
  paste0(cohort, "_fdr")
)

res_withInfo_df <- cbind(info_df, res_df)

saveRDS(
  res_withInfo_df,
  paste0(data.dir.median, cohort, "_linear_df.rds")
)
```

```{R}
file <- dir(data.dir.median,pattern = paste0(cohort,"_linear_df"),
            recursive = T,
            full.names = TRUE,
            ignore.case = T)
file
res_withInfo_df <- readRDS(file)
dim(res_withInfo_df)
res_withInfo_df
```

# DMRs Validation

```{R, eval = FALSE}
### Call in datasets
main_dmrs <- read.csv(
    "meta_analysis_region_results/step4_dmr_vs_cpgs/meta_analysis_sig_no_crossHyb_smoking_ov_comb_p_with_sig_single_cpgs.csv"
) #dim:119 60
blood_dmrs <- readRDS(
  dir(data.dir.median,pattern = "London_blood_linear_df.RDS",ignore.case = TRUE,full.names = T)
) #dim: 40104 8

### select the most significant cluster for each input region 
london_blood_selected <- blood_dmrs %>%
  group_by(LONDON_blood_inputRegion) %>%
  filter(LONDON_blood_pVal == min(LONDON_blood_pVal))


### select the most significant cluster for each input region  
london_blood_metaSig <- london_blood_selected[
  match(
    main_dmrs$inputRegion,
    london_blood_selected$LONDON_blood_inputRegion
  ),
]
london_blood_metaSig <- london_blood_metaSig[
  !is.na(london_blood_metaSig$LONDON_blood_inputRegion),
]
dim(london_blood_metaSig)
# [1] 119   8

### calculate validation fdr
london_blood_metaSig$LONDON_blood_validation_fdr <- p.adjust(
  london_blood_metaSig$LONDON_blood_pVal, method = "fdr"
)

### add annotation
london_blood_metaSig_withAnnot <- merge(
  main_dmrs[,1:5], 
  london_blood_metaSig,
  by.x = "inputRegion",
  by.y = "LONDON_blood_inputRegion",
  sort = FALSE
)
write.csv(
  london_blood_metaSig_withAnnot,
  paste0(data.dir.validation, "validation_london_blood.csv"),
  row.names = FALSE
)
```

# Linear regression by single cpgs 

```{R, eval = TRUE}
### Import datasets
beta_mat <- readRDS(
    paste0(data.dir.pca, "London_QNBMIQ_PCfiltered.RDS")
) 
pheno_df <- readRDS(
 dir(data.dir.neuron,
     pattern =  "pheno_BLOOD_withBloodProp_df.RDS",
     ignore.case = TRUE,
     full.names = TRUE)
)

pheno_df$Sample <- pheno_df$sample

identical(pheno_df$Sample, colnames(beta_mat))

### Compute M values
mval_mat <- log2(beta_mat / (1 - beta_mat))

pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)

str(pheno_df)

is(pheno_df$age.blood, "numeric")
```

```{R, eval = FALSE}
predictors_char <- "status"
covariates_char <- c(
  "age.blood", "sex", "slide",
  "B", "NK", "CD4T", "CD8T", "Mono", "Neutro", "Eosino"
)

doParallel::registerDoParallel(cores = parallel::detectCores()/2)
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

results_ordered_df <- plyr::adply(mval_mat,1, function(row){
  
  sumOneRegion_df <- as.data.frame(t(row))
  
  result <- TestSingleRegion(
    predictors_char = predictors_char,
    covariates_char = covariates_char,
    pheno_df = pheno_df,
    sumOneRegion_df = sumOneRegion_df
  )
  result
},
.progress = "time",
.parallel = TRUE,
.id = "cpg")
colnames(results_ordered_df)[1] <- "cpg"

identical(row.names(mval_mat), results_ordered_df$cpg %>% as.character())

write.csv(
  results_ordered_df,
  paste0(data.dir.median, "London_blood_single_cpg_linear_df.csv"),
  row.names = FALSE
)
```

```{R}
results_ordered_df <- readr::read_csv(
    paste0(data.dir.median, "London_blood_single_cpg_linear_df.csv")
)
results_ordered_df
```

# CpGs Validation

```{R, eval = FALSE}
### Call in datasets
main_cpgs <- read.csv(
    "meta_analysis_single_cpg_results/meta_analysis_single_cpg_sig_no_crossHyb_smoking_df.csv"
) #dim:3751 28
blood_cpgs <- read.csv(
    paste0(data.dir.median, "London_blood_single_cpg_linear_df.csv")
) #dim: 450793 4

### subset validation datasets with significant cpgs in meta analysis
blood_metaSig <- blood_cpgs[
  match(
      as.character(main_cpgs$cpg),
      as.character(blood_cpgs$cpg)
  ),
]
blood_metaSig <- blood_metaSig[
  !is.na(blood_metaSig$cpg),
]
dim(blood_metaSig)
# [1] 3751    4

### calculate validation fdr
blood_metaSig$validation_fdr <- p.adjust(
  blood_metaSig$pValue, method = "fdr"
)

### add annotation
blood_metaSig_withAnnot <- merge(
  main_cpgs[,1:8], blood_metaSig,
  by.x = "cpg",
  by.y = "cpg",
  sort = FALSE
)

write.csv(
  blood_metaSig_withAnnot,
  paste0(data.dir.validation, "validation_single_cpg_blood.csv"),
  row.names = FALSE
)
```

# Data final

```{R}
dir(path = data.dir,recursive = T,pattern = ".rda|.csv|.RDS")
```

# Session information
```{R}
devtools::session_info()
```
