---
title: "LONDON blood dataset"
author: "Lanyu Zhang, Tiago C. Silva, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```


# Data retrival

```{R, message = FALSE, results = 'hide'}
library(dplyr)
```

```{R}
cohort <- "LONDON_blood"
data.dir <- file.path("DATASETS/",cohort,"/") 
data.dir.table <- "DATASETS/Summary_Table/" 
data.dir.raw <- file.path(data.dir,"/step1_download/") 
data.dir.clinical.filter <- file.path(data.dir,"/step2_clinical_available_filtering/") 
data.dir.probes.qc <- file.path(data.dir,"/step3_probesQC_filtering/") 
data.dir.probes.normalization <- file.path(data.dir,"/step4_normalization/") 
data.dir.pca <- file.path(data.dir,"/step5_pca_filtering/") 
data.dir.neuron <- file.path(data.dir,"/step6_neuron_comp/") 
# data.dir.single.cpg.pval <- file.path(data.dir,"/step7_single_cpg_pval/") 
data.dir.residuals <- file.path(data.dir,"/step7_residuals/") 
data.dir.median <- file.path(data.dir,"/step8_median/")
data.dir.validation <- file.path(data.dir,"/step9_validation/") 
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

Required R library `GEOquery` can be installed as following:

```{R, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("GEOquery")
```

Source: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE59685

```{r synapse, eval = FALSE}
library(GEOquery)
library(SummarizedExperiment)
test <- getGEO(GEO = "GSE59685",destdir = data.dir,GSEMatrix = TRUE,)
GSE59685 <- test$GSE59685_series_matrix.txt.gz %>% makeSummarizedExperimentFromExpressionSet() 
metadata <- colData(GSE59685)
## Failed to create assayData with pacakage GEOquery, so we use the beta matrix
## directly downloaded online instead
library(data.table)
assayData <- fread(
  "ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE59nnn/GSE59685/suppl/GSE59685_betas.csv.gz",
  skip = 5,
  header = TRUE,
  sep =','
)

## Turn assayData from "data.table" and "data.frame" to "data.frame" only
assayData <- data.frame(assayData)

## Exclude row 1 as it's old sample id
assayData <- assayData[-1, ]

## Create row names based on V1
row.names(assayData) <- assayData$V1

## Delete column V1 after creating rownames
assayData <- assayData[, 2:ncol(assayData)]

assayData <- data.matrix(assayData)

## Save RDS file
saveRDS(assayData, paste0(data.dir.raw, "GSE59685_assay.RDS"))

### Extract and save phenotype dataset
phenoData <-  colData(GSE59685)
phenoData <- phenoData[match(colnames(assayData),rownames(phenoData)),]
saveRDS(phenoData, paste0(data.dir.raw, "GSE59685_pheno.RDS"))
write.csv(phenoData, paste0(data.dir.raw, "GSE59685_pheno.csv"))
```

#  Data Pre-processing 

Description:

 Subset Files
 1. subset pheno data to
  1) whole blood tissue 
  2) variables -- sample, subject_id, sentrix_id, slide, age, sex, stage
 2. subset methylation data to match subjects in pheno data

 Input: GSE59685_pheno.csv, GSE59685_assay.RDS
 Output: pheno_BLOOD_df.RDS, beta110_BLOOD_mat.RDS



```{R, eval = TRUE}
##### 1. Subset pheno data #####################################################

### Read in phenotype data
phenoRaw_df <- readr::read_csv(paste0(data.dir.raw, "GSE59685_pheno.csv")) #dim: 531 49

### Subset rows and columns

### Subset rows and columns
phenoBLOOD_df <- phenoRaw_df[
  (phenoRaw_df$source_name_ch1 == "whole blood"),
  c("geo_accession", "subjectid.ch1", "barcode.ch1",
    "age.blood.ch1", "Sex.ch1", "ad.disease.status.ch1", "braak.stage.ch1")
]

### Rename vars
colnames(phenoBLOOD_df) <- c(
  "sample", "subject.id", "sentrix_id", "age.blood", "sex", "status", "stage"
)

### Turn factors into characters
phenoBLOOD_df$sample <- as.character(phenoBLOOD_df$sample)
phenoBLOOD_df$subject.id <- as.character(phenoBLOOD_df$subject.id)
phenoBLOOD_df$sentrix_id <- as.character(phenoBLOOD_df$sentrix_id)
phenoBLOOD_df$sex <- as.character(phenoBLOOD_df$sex)
phenoBLOOD_df$status <- as.character(phenoBLOOD_df$status)
phenoBLOOD_df$stage <- as.integer(as.character(phenoBLOOD_df$stage))

### Get slide from sentrix_id
  # e.g. "6042316048_R05C01"(sentrix_id) -- "6042316048"(slide) and "R05C01"(array)
sentrixID_mat <- do.call(rbind, strsplit(phenoBLOOD_df$sentrix_id, "_"))
phenoBLOOD_df$slide <- sentrixID_mat[, 1]

### Order final pheno_df
pheno_df <- phenoBLOOD_df[, c(
  "sample", "subject.id", "sentrix_id", "slide",
  "age.blood", "sex", "status", "stage"
)] 
dim(pheno_df)
```

```{R}
##### 2. Subset methylation data ###############################################

### Read in methylation data
beta_mat <- readRDS(paste0(data.dir.raw, "GSE59685_assay.RDS")) #dim: 485577 531

### Subset methylation data based on pheno data
beta_mat <- beta_mat[, match(pheno_df$sample,colnames(beta_mat))] # dim: 485577 110
```

```{R, eval = FALSE}
##### 3. Output datasets #######################################################

## phenotype dataset
saveRDS(pheno_df, paste0(data.dir.clinical.filter, "pheno_BLOOD_withStatusExclude_df.RDS")) 

## methylation beta values dataset
saveRDS(beta_mat, paste0(data.dir.clinical.filter, "beta_BLOOD_withStatusExclude_mat.RDS")) 
```

## Probes QC

 1. keep only probes that start with "cg"
 2. drop probes that are on X/Y
 3. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.

 Input: beta110_PFC_mat.RDS
 
 Output: beta110_PFC_CG_XY_SNPfiltered_mat.RDS

```{R}
##### 1. keep on probes with start with "cg" ###################################
beta_mat <- readRDS(paste0(data.dir.clinical.filter, "beta_BLOOD_withStatusExclude_mat.RDS")) 
nb.probes <- nrow(beta_mat)
nb.samples <- ncol(beta_mat)
nb.samples.with.clinical <- ncol(beta_mat)

beta_mat <- beta_mat[grep("cg",rownames(beta_mat)),]
dim(beta_mat)
nb.probes.cg <- nrow(beta_mat)
```

```{R, message = FALSE, results = "hide"}
##### 2. drop probes that are on X/Y ###########################################
##### 3. drop probes where SNP with MAF >= 0.01 in the last 5 bp of the probe ##
library(DMRcate)
```

```{R, message = FALSE, warning = FALSE}
beta_CG_XY_SNPfiltered_mat <- rmSNPandCH(
  object = beta_mat,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = FALSE,
  rmXY = TRUE
)
dim(beta_CG_XY_SNPfiltered_mat)
nb.probes.cg.dmrcate <- nrow(beta_CG_XY_SNPfiltered_mat)
##### 4. Output datasets #######################################################

saveRDS(
  beta_CG_XY_SNPfiltered_mat,
  paste0(data.dir.probes.qc, "beta_CG_XY_SNPfiltered_withStatusExclude_mat.RDS")
)

```

## Samples QC

- Quantile normalization and BMIQ normalization

 Input: 
 
 - beta_CG_XY_SNPfiltered_mat.RDS
 - pheno_PFC_df.RDS
 - full.annot.RDS
 
 Output: 
 
- London_PFC_QNBMIQ.RDS

### Quantile normalization

```{R, eval = FALSE}
library(lumi)
betaQN <- lumiN(x.lumi = beta_mat, method = "quantile")

dim(betaQN)
```

```{R, message = FALSE, results = "hide"}
##### 5. BMIQ ##################################################################
library(wateRmelon)
library(RPMM)
library(sesame)
library(sesameData)
```

```{R, eval = FALSE}
### Order annotation in the same order as beta matrix
annotType <- sesameDataGet("HM450.hg19.manifest")
annotType$designTypeNumeric <- ifelse(annotType$designType == "I",1,2)

### Density plot for type I and type II probes
library(sm)

betaQNCompleteCol1 <- betaQN[complete.cases(betaQN[,1]), ]
annotTypeCompleteCol1 <- annotType[row.names(betaQNCompleteCol1), ]

sm.density.compare(
  betaQNCompleteCol1[,1],
  annotTypeCompleteCol1$designTypeNumeric
)

type12 <- annotType$designTypeNumeric[match(rownames(betaQN),names(annotType))]
```

```{R, eval = FALSE}
### BMIQ
set.seed (946)
doParallel::registerDoParallel(cores = 8)
betaQN_BMIQ <- plyr::aaply(
    betaQN, 2,
    function(x){
        norm_ls <- BMIQ(x, design.v = type12, plots = FALSE)
        return (norm_ls$nbeta)
    },.progress = "time",.parallel = TRUE
) %>% t()

saveRDS(betaQN_BMIQ, paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ_withStatusExclude.RDS"))
```


```{R, include = FALSE}
betaQN_BMIQ <- readRDS(paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ_withStatusExclude.RDS"))
```


# Outliers detection - PCA analysis


Description: 

1. estimate standard deviation for each probe
2. select most variable probes (e.g. n = 50,000)
3. pca plot
4. Filter outliers

Input: 

- QNBMIQ.rds
- pheno_df.RDS

Output: 

- PCs_usingBetas.csv, 
- PCA plots
- QNBMIQ_PCfiltered.RDS
- pheno_df.RDS


```{R}
# plotPCA and OrderDataBySd functions
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

beta_mat <- readRDS(paste0(data.dir.probes.normalization, "London_BLOOD_QNBMIQ_withStatusExclude.RDS")) #dim: 437713 110

pheno_df <- readRDS(paste0(data.dir.clinical.filter, "pheno_BLOOD_withStatusExclude_df.RDS")) #dim: 110 7

identical(colnames(beta_mat), pheno_df$sample)

### transform to m values
mvalue_mat <- log2(beta_mat/(1 - beta_mat)) #dim: 437713 110

pheno_df <- subset(pheno_df, pheno_df$sample %in% colnames(beta_mat)) #dim: 110 7
##### 1.Order matrix by most variable probes on top ############################

betaOrd_mat <- OrderDataBySd(beta_mat) #dim: 437713 110

mOrd_mat <- OrderDataBySd(mvalue_mat)  #dim: 437713 110

betaOrd_matPlot <- betaOrd_mat[, pheno_df$sample] #dim: 437713 110
mOrd_matPlot <- mOrd_mat[, pheno_df$sample]       #dim: 437713 110
identical(pheno_df$sample, colnames(betaOrd_matPlot))
identical(pheno_df$sample, colnames(mOrd_matPlot))

expSorted_mat = betaOrd_mat #dim: 437713 110

pca <- prcomp(
  t(expSorted_mat[1:50000,]),
  center = TRUE,
  scale = TRUE
)


d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2])

meanPC1 <- mean (d$PC1)
sdPC1   <- sd (d$PC1)

meanPC2 <- mean (d$PC2)
sdPC2   <- sd (d$PC2)

out3sdPC1_1 <- meanPC1 - 3*sdPC1
out3sdPC1_2 <- meanPC1 + 3*sdPC1

out3sdPC2_1 <- meanPC2 - 3*sdPC2
out3sdPC2_2 <- meanPC2 + 3*sdPC2


d$outlier_PC1[d$PC1 >= out3sdPC1_1 & d$PC1 <= out3sdPC1_2] <- 0
d$outlier_PC1[d$PC1 < out3sdPC1_1 | d$PC1 > out3sdPC1_2] <- 1

d$outlier_PC2[d$PC2 >= out3sdPC2_1 & d$PC2 <= out3sdPC2_2] <- 0
d$outlier_PC2[d$PC2 < out3sdPC2_1 | d$PC2 > out3sdPC2_2] <- 1

write.csv(d, paste0(data.dir.pca, "London_Blood_PCs_usingBetas_withStatusExclude.csv"))


##### 2.PCA plot ###############################################################
library(ggplot2)
library(ggrepel)

### beta values
byStatus <- plotPCA(
  dataset = "London Blood: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "status",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "London Blood: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

### M values
byStatus <- plotPCA(
  dataset = "London Blood: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "status",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "London Blood: M values",
  expSorted_mat = mOrd_mat,
  pheno = pheno_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
```

## Filter samples by PCA 
```{R, eval = FALSE}
noOutliers <- d[which(d$outlier_PC1 == 0 & d$outlier_PC2 == 0), ]
betaQN_BMIQ_PCfiltered <- beta_mat[, rownames(noOutliers)]
saveRDS(betaQN_BMIQ_PCfiltered, paste0(data.dir.pca, "London_QNBMIQ_PCfiltered_withStatusExclude.RDS"))

pheno_df <- pheno_df[pheno_df$sample %in% rownames(noOutliers),] 
saveRDS(pheno_df, paste0(data.dir.pca, "pheno_withStatusExclude_df.RDS"))
```

# Summary after QC steps

## Data and metadata
```{R}
betaQN_BMIQ_PCfiltered <- readRDS(paste0(data.dir.pca, "London_QNBMIQ_PCfiltered_withStatusExclude.RDS")) 
nb.samples.with.clinical.after.pca <- ncol(betaQN_BMIQ_PCfiltered)
pheno_df <- readRDS(paste0(data.dir.pca, "pheno_withStatusExclude_df.RDS"))
dim(betaQN_BMIQ_PCfiltered)
dim(pheno_df)
```

```{R}
pheno_df %>% 
  DT::datatable(filter = 'top',
                style = "bootstrap",
                extensions = 'Buttons',
                options = list(scrollX = TRUE, 
                               dom = 'Bfrtip',
                               buttons = I('colvis'),
                               keys = TRUE, 
                               pageLength = 10), 
                rownames = FALSE,
                caption = "Samples metadata")
```



## Numbers of samples and probes removed in each step


```{R}
df.samples <- data.frame("Number of samples" =  c(nb.samples, 
                                                  nb.samples.with.clinical, 
                                                  nb.samples.with.clinical.after.pca),
                         "Description" = c("total number of samples",
                                           "samples with clinical data",
                                           "Samples after PCA"),
                         "Difference" = c("-",
                                          nb.samples.with.clinical - nb.samples ,
                                          nb.samples.with.clinical.after.pca - nb.samples.with.clinical)
)    
df.samples                     
# Create summary table
df.probes <- data.frame("Number of probes" = c(nb.probes,
                                               nb.probes.cg, 
                                               nb.probes.cg.dmrcate),
                        "Description" = c("total number of probes in raw data",
                                          "only probes that start with cg",
                                          "DMRcate"),
                        "Difference" = c("-",
                                         nb.probes.cg - nb.probes ,
                                         nb.probes.cg.dmrcate - nb.probes.cg)
)
df.probes
save(df.samples,df.probes,file = file.path(data.dir.table, "LONDON_blood_table.rda"))
```

# Compute cell type proportions 


Data from  https://www.tandfonline.com/doi/full/10.4161/epi.23924

- Input: London_QNBMIQ_PCfiltered.RDS, pheno_blood_df.RDS
- Output: pheno_BLOOD_withBloodProp_df.rds

```{R, FALSE}
blood <- readRDS(paste0(data.dir.pca, "London_QNBMIQ_PCfiltered_withStatusExclude.RDS")) 
nb.samples.with.clinical.after.pca <- ncol(blood)
pheno <- readRDS(paste0(data.dir.pca, "pheno_withStatusExclude_df.RDS"))
library(EpiDISH)

data(centDHSbloodDMC.m)

out.l <- epidish(blood, centDHSbloodDMC.m, method = 'RPC')
frac.m <- data.frame(out.l$estF)

pheno_final <- merge(
  pheno, 
  frac.m,
  by.x = "sample",
  by.y = "row.names",
  sort = FALSE
)
identical(pheno_final$sample, colnames(blood))

saveRDS(
  pheno_final,
  paste0(data.dir.neuron, "pheno_BLOOD_withBloodProp_withStatusExclude_df.rds")
)
```

# Session information
```{R}
devtools::session_info()
```
