---
title: "MtSinai dataset"
author: "Lanyu Zhang, Tiago C. Silva, Lily Wang"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
    theme: lumen
    toc: true
    number_sections: true
    df_print: paged
    code_download: false
    toc_float:
      collapsed: yes
    toc_depth: 3
editor_options:
  chunk_output_type: inline    
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,warning = FALSE)
```


# Data retrival

```{R, message = FALSE, results = 'hide'}
library(dplyr)
```

```{R}
cohort <- "MtSinai"
data.dir <- file.path("DATASETS/",cohort,"/") 
data.dir.table <- "DATASETS/Summary_Table/" 
data.dir.raw <- file.path(data.dir,"/step1_download/") 
data.dir.clinical.filter <- file.path(data.dir,"/step2_clinical_available_filtering/") 
data.dir.probes.qc <- file.path(data.dir,"/step3_probesQC_filtering/") 
data.dir.probes.normalization <- file.path(data.dir,"/step4_normalization/") 
data.dir.pca <- file.path(data.dir,"/step5_pca_filtering/") 
data.dir.neuron <- file.path(data.dir,"/step6_neuron_comp/") 
data.dir.single.cpg.pval <- file.path(data.dir,"/step7_single_cpg_pval/") 
data.dir.residuals <- file.path(data.dir,"/step10_residuals/") 
data.dir.median <- file.path(data.dir,"/step11_median/") 
for(p in grep("dir",ls(),value = T)) dir.create(get(p),recursive = TRUE,showWarnings = FALSE)
```

Required R library `GEOquery` can be installed as following:

```{R, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install("GEOquery")
```

Data from: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE80970

```{r synapse, eval = FALSE}
library(GEOquery)
dir.create(data.dir,recursive = TRUE,showWarnings = FALSE)
test <- getGEO(GEO = "GSE80970",destdir = data.dir.raw)

assayData <- exprs(test$GSE80970_series_matrix.txt.gz)
saveRDS(assayData, paste0(data.dir.raw, "GSE80970_assay.RDS"))
write.csv(assayData, paste0(data.dir.raw, "GSE80970_assay.csv"))

phenoData <- pData(test$GSE80970_series_matrix.txt.gz)
saveRDS(phenoData, paste0(data.dir.raw, "GSE80970_pheno.RDS"))
write.csv(phenoData, paste0(data.dir.raw, "GSE80970_pheno.csv"))
```

# Data Pre-processing

## Subset samples

Subset Files:

- subset pheno data to
  1. prefrontal cortex tissue (n = 142)
  2. variables -- sample, subject_id, sentrix_id, slide, age, sex, stage
- subset methylation data to match subjects in pheno data (n = 142)

- Input: GSE80970_pheno.csv, GSE80970_assay.RDS

- Output: pheno_df.RDS, beta_mat.RDS

```{r data_procesing_libs}
### Read in phenotype data
phenoRaw_df <- readr::read_csv( paste0(data.dir.raw, "GSE80970_pheno.csv"),
                                col_types = readr::cols())
colnames(phenoRaw_df) <- gsub(" |[[:punct:]]",".",colnames(phenoRaw_df))
dim(phenoRaw_df)

### Subset rows and columns
phenoPFC_df <- phenoRaw_df %>% 
  dplyr::filter(tissue.ch1 == "prefrontal cortex") %>%
  dplyr::select(c("geo.accession", 
                  "donor.id.ch1",
                  "sentrix.id.ch1",
                  "age..yr..ch1",
                  "gender.ch1",
                  "braak.stage.ch1"))

### Rename vars
colnames(phenoPFC_df) <- c(
  "sample", "subject.id", "sentrix_id", "age.brain", "sex", "stage"
)

### Get slide from sentrix_id
# e.g. "7796806144_R01C01"(sentrix_id) -- "7796806144"(slide) and "R01C01"(array)
phenoPFC_df$slide <- gsub("_[[:alnum:]]*$","",phenoPFC_df$sentrix_id)

### Order final pheno_df
pheno_df <- phenoPFC_df[, c(
  "sample", "subject.id", "sentrix_id", "slide", "age.brain", "sex", "stage"
)] 
```

```{R}
##### 2. Subset methylation data ###############################################

### Read in methylation data
beta_mat <- readRDS(paste0(data.dir.raw, "GSE80970_assay.RDS"))

### Subset methylation data based on pheno data
beta_mat <- beta_mat[, pheno_df$sample] 
all(colnames(beta_mat) ==  pheno_df$sample)
nb.probes <- nrow(beta_mat)
nb.samples <- ncol(beta_mat)
nb.samples.with.clinical <- ncol(beta_mat)
```

```{R, eval = FALSE}
##### 3. Output datasets #######################################################
## phenotype dataset
saveRDS(pheno_df, paste0(data.dir.clinical.filter, "pheno_df.RDS"))

## methylation beta values dataset
saveRDS(beta_mat, paste0(data.dir.clinical.filter, "beta142_mat.RDS"))
```


## QC probes

1. keep only probes that start with "cg"
2. drop probes that are on X/Y
3. drop probes where SNP with MAF >= 0.01 was present in the last 5 bp of the probe.

Input: beta_mat.RDS
Output: beta_CG_XY_SNPfiltered_mat.RDS

```{R, message = FALSE, results = "hide", include = FALSE}
library(sesameData)  
```

```{R}
##### 1. keep on probes with start with "cg" ###################################
beta142_mat <- readRDS(paste0(data.dir.clinical.filter, "beta142_mat.RDS")) #dim: 485577 142
dim(beta142_mat)

beta142cg_mat <- beta142_mat[grep("cg",rownames(beta142_mat)),]
dim(beta142cg_mat)
nb.cg.probes <- nrow(beta142cg_mat)
```

```{R, message = FALSE, results = "hide"}
##### 2. drop probes that are on X/Y ###########################################
##### 3. drop probes where SNP with MAF >= 0.01 in the last 5 bp of the probe ##
library(DMRcate)
```

```{R, message = FALSE, warning = FALSE}
beta142Filtered_mat <- rmSNPandCH(
  object = beta142cg_mat,
  dist = 5,
  mafcut = 0.01,
  and = TRUE,
  rmcrosshyb = FALSE,
  rmXY = TRUE
) #dim: 437713 142
dim(beta142Filtered_mat)
nb.probes.cg.dmrcate <- nrow(beta142Filtered_mat)
```

```{R, eval = TRUE}
##### 4. Output datasets #######################################################
saveRDS(
  beta142Filtered_mat,
  paste0(data.dir.probes.qc, "beta142_CG_XY_SNPfiltered_mat.RDS")
)
```

## QC samples

- Quantile normalization and BMIQ normalization

Input: 

- beta_CG_XY_SNPfiltered_mat.RDS
- pheno_df.RDS
- full.annot.RDS

Output: 

- MtSinai_QNBMIQ.RDS

```{R qcSamplesData}
beta_mat <- readRDS(paste0(data.dir.probes.qc, "beta142_CG_XY_SNPfiltered_mat.RDS"))
pheno_df <- readRDS(paste0(data.dir.clinical.filter, "pheno_df.RDS"))
```

## Quantile normalization 

```{R, eval = FALSE}
library(lumi)
betaQN <- lumiN(x.lumi = beta_mat, method = "quantile")
```

```{R, message = FALSE, results = "hide"}
##### 5. BMIQ ##################################################################
library(wateRmelon)
library(RPMM)
library(sesame)
library(sesameData)
```

```{R, eval = FALSE}
### Order annotation in the same order as beta matrix
annotType <- sesameDataGet("HM450.hg19.manifest")
annotType$designTypeNumeric <- ifelse(annotType$designType == "I",1,2)

### Density plot for type I and type II probes
library(sm)

betaQNCompleteCol1 <- betaQN[complete.cases(betaQN[,1]), ]
annotTypeCompleteCol1 <- annotType[row.names(betaQNCompleteCol1), ]

sm.density.compare(
  betaQNCompleteCol1[,1],
  annotTypeCompleteCol1$designTypeNumeric
)

type12 <- annotType$designTypeNumeric[match(rownames(betaQN),names(annotType))]
```

```{R, eval = FALSE}
### BMIQ
set.seed (946)
betaQN_BMIQ <- apply(
  betaQN, 2,
  function(x){
    norm_ls <- BMIQ(x, design.v = type12, plots = FALSE)
    return (norm_ls$nbeta)
  }
)
saveRDS(betaQN_BMIQ, paste0(data.dir.probes.normalization, "MtSinai_QNBMIQ.RDS"))
```

```{R, include = FALSE}
betaQN_BMIQ <- readRDS(paste0(data.dir.probes.normalization, "MtSinai_QNBMIQ.RDS"))
```

# Outliers detection - PCA analysis

Description: 

1. estimate standard deviation for each probe
2. select most variable probes (e.g. n = 50,000)
3. PCA analysis
4. Filter outliers

Input: 

- MtSinai_QNBMIQ.rds
- pheno_df.RDS

Output: 

- PCs_usingBetas.csv, 
- PCA plots
- QNBMIQ_PCfiltered.RDS
- pheno_df.RDS


```{R}
# plotPCA and OrderDataBySd functions
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")
```

```{R}
##### 0.Import datasets ########################################################
beta_mat <- readRDS(paste0(data.dir.probes.normalization, "MtSinai_QNBMIQ.RDS")) #dim: 437713 142

pheno_df <- readRDS(paste0(data.dir.clinical.filter, "pheno_df.RDS")) #dim: 142 7

identical(colnames(beta_mat), pheno_df$sample)

### transform to m values
mvalue_mat <- log2(beta_mat / (1 - beta_mat)) #dim: 437713 142

both_df <- subset(pheno_df, pheno_df$sample %in% colnames(beta_mat)) #dim: 142 7

# both_df$stage <- as.numeric(as.character(both_df$stage))

both_df$stage3 <- both_df$stage
both_df$stage3[both_df$stage <= 2] <- '0-2'
both_df$stage3[both_df$stage > 2 & both_df$stage < 5] <- '3-4'
both_df$stage3[both_df$stage >= 5] <- '5-6'


##### 1.Order matrix by most variable probes on top ############################

betaOrd_mat <- OrderDataBySd(beta_mat) #dim: 391482 142

mOrd_mat <- OrderDataBySd(mvalue_mat)  #dim: 391482 142

betaOrd_matPlot <- betaOrd_mat[, both_df$sample] #dim: 391482 142
mOrd_matPlot <- mOrd_mat[, both_df$sample]       #dim: 391482 142
identical(both_df$sample, colnames(betaOrd_matPlot))
identical(both_df$sample, colnames(mOrd_matPlot))

expSorted_mat = betaOrd_mat #dim: 391482 142

pca <- prcomp(t(expSorted_mat[1:50000, ]),
              center = TRUE,
              scale = TRUE)

d <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2])

meanPC1 <- mean (d$PC1)
sdPC1   <- sd (d$PC1)

meanPC2 <- mean (d$PC2)
sdPC2   <- sd (d$PC2)

out3sdPC1_1 <- meanPC1 - 3 * sdPC1
out3sdPC1_2 <- meanPC1 + 3 * sdPC1

out3sdPC2_1 <- meanPC2 - 3 * sdPC2
out3sdPC2_2 <- meanPC2 + 3 * sdPC2

d$outlier_PC1[d$PC1 >= out3sdPC1_1 & d$PC1 <= out3sdPC1_2] <- 0
d$outlier_PC1[d$PC1 < out3sdPC1_1 | d$PC1 > out3sdPC1_2] <- 1

d$outlier_PC2[d$PC2 >= out3sdPC2_1 & d$PC2 <= out3sdPC2_2] <- 0
d$outlier_PC2[d$PC2 < out3sdPC2_1 | d$PC2 > out3sdPC2_2] <- 1

readr::write_csv(d, paste0(data.dir.pca, "MtSinai_PCs_usingBetas.csv"))
```

```{R, include = FALSE, eval = FALSE}

##### 2.PCA plot ###############################################################

library(ggplot2)
library(ggrepel)

### beta values
byStage <- plotPCA(
  dataset = "MtSinai: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = both_df,
  group_char = "stage3",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "MtSinai: beta values",
  expSorted_mat = betaOrd_mat,
  pheno = both_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)


### M values
byStage <- plotPCA(
  dataset = "MtSinai: M values",
  expSorted_mat = mOrd_mat,
  pheno = both_df,
  group_char = "stage3",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)

bySex <- plotPCA(
  dataset = "MtSinai: M values",
  expSorted_mat = mOrd_mat,
  pheno = both_df,
  group_char = "sex",
  ntop = 50000,
  center = TRUE,
  scale = TRUE
)
```


```{R}
##### 3.Filter samples by PCA, save files ######################################
noOutliers <- d[which(d$outlier_PC1 == 0 & d$outlier_PC2 == 0), ]
betaQN_BMIQ_PCfiltered <- beta_mat[, rownames(noOutliers)] #dim: 437713 141
saveRDS(betaQN_BMIQ_PCfiltered, paste0(data.dir.pca, "MtSinai_QNBMIQ_PCfiltered.RDS"))

pheno141_df <- pheno_df[pheno_df$sample %in% rownames(noOutliers),] #dim: 141 7
saveRDS(pheno141_df, paste0(data.dir.pca, "pheno141_df.RDS"))
```

# Summary after QC steps

## Data and metadata
```{R}
nb.samples.with.clinical.after.pca <- ncol(betaQN_BMIQ_PCfiltered)

dim(betaQN_BMIQ_PCfiltered)
dim(pheno141_df)
pheno141_df %>% 
  DT::datatable(filter = 'top',
                style = "bootstrap",
                extensions = 'Buttons',
                options = list(scrollX = TRUE, 
                               dom = 'Bfrtip',
                               buttons = I('colvis'),
                               keys = TRUE, 
                               pageLength = 10), 
                rownames = FALSE,
                caption = "Samples metadata")
```


## Numbers of samples and probes removed in each step


```{R}
df.samples <- data.frame("Number of samples" =  c(nb.samples, 
                                                  nb.samples.with.clinical, 
                                                  nb.samples.with.clinical.after.pca),
                         "Description" = c("total number of samples",
                                           "samples with clinical data",
                                           "Samples after PCA"),
                         "Difference" = c("-",
                                          nb.samples.with.clinical - nb.samples ,
                                          nb.samples.with.clinical.after.pca - nb.samples.with.clinical)
)    
df.samples                     
# Create summary table
df.probes <- data.frame("Number of probes" = c(nb.probes,
                                               nb.cg.probes, 
                                               nb.probes.cg.dmrcate),
                        "Description" = c("total number of probes in raw data",
                                          "only probes that start with cg",
                                          "DMRcate"),
                        "Difference" = c("-",
                                         nb.cg.probes - nb.probes ,
                                         nb.probes.cg.dmrcate - nb.cg.probes)
)
df.probes
save(df.samples,df.probes,file = file.path(data.dir.table, "MtSiani_table.rda"))
```

# Compute neuron proportion


Data from  https://www.tandfonline.com/doi/full/10.4161/epi.23924

- Input: MtSinai_QNBMIQ_PCfiltered.RDS, pheno141_df.RDS
- Output: pheno141_withNeuronProp_df.RDS

```{R}
objects <- load("../../CET/CETS_Image.RData")
objects
```

## 1. Get reference profile from Caucasions + controls 
```{R}
idx <- list(
  controlNeuron = pdBrain$celltype == "N" & pdBrain$diag == "Control" & pdBrain$ethnicity == "Caucasian",
  controlGlia   = pdBrain$celltype == "G" & pdBrain$diag == "Control" & pdBrain$ethnicity == "Caucasian"
)

refProfile <- getReference(brain, idx)


##### 2. Estimate proportions of neurons in PFC samples ########################

### Limit to 10,000 cpgs in the refProfile dataset
pfc <- readRDS(paste0(data.dir.pca, "MtSinai_QNBMIQ_PCfiltered.RDS")) #dim: 433656 59

selected <- rownames(pfc) %in% rownames(refProfile)

pfc.refcpgs <- pfc[selected, ] #dim: 9530 59

### Estimate proportion of neurons
prop <- data.frame(estProportion(pfc.refcpgs, profile = refProfile))
colnames(prop) <- "prop.neuron"


##### 3. Merge pfc.refcpgs with phenotype file #################################
pheno <- readRDS(paste0(data.dir.pca, "pheno141_df.RDS"))

pheno_final <- merge(
  pheno,
  prop,
  by.x = "sample",
  by.y = "row.names"
)

saveRDS(pheno_final, paste0(data.dir.neuron, "pheno141_withNeuronProp_df.RDS"))
```

# Linear regression by cpgs Methylation 

Input: 

- QNBMIQ_PCfiltered.RDS,
- pheno_withNeuronProp_df.RDS

Output:

- single_cpg_pVal_df.csv

## 1. Import datasets

```{R}
beta_mat <- readRDS(paste0(data.dir.pca, "MtSinai_QNBMIQ_PCfiltered.RDS")) #dim:433656 59
pheno_df <- readRDS(paste0(data.dir.neuron, "pheno141_withNeuronProp_df.RDS")) #dim:59 7
```

## 2. Test all regions

```{R, eval = TRUE}
### Compute M values
mval_mat <- log2(beta_mat / (1 - beta_mat))

pheno_df$Sample <- pheno_df$sample

identical(pheno_df$Sample, colnames(mval_mat))

pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)
# If rosmap cohort, don't forget batch effect

str(pheno_df)

is(pheno_df$stage,"numeric")
is(pheno_df$age.brain,"numeric")
is(pheno_df$prop.neuron,"numeric")
```

```{R, eval = FALSE}

predictors_char <- "stage"
covariates_char <- c("age.brain", "sex", "prop.neuron", "slide")

doParallel::registerDoParallel(cores = parallel::detectCores()/2)
devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

results_ordered_df <- plyr::adply(mval_mat,1, function(row){
  
  sumOneRegion_df <- data.frame(t(row))
  
  result <- TestSingleRegion(
    predictors_char = predictors_char,
    covariates_char = covariates_char,
    pheno_df = pheno_df,
    sumOneRegion_df = sumOneRegion_df
  )
  result
}, .progress = "time",.parallel = T,.id = "cpg")
colnames(results_ordered_df)[1] <- "cpg"  


identical(row.names(mval_mat), results_ordered_df$cpg %>% as.character())

results_ordered_df$fdr <- p.adjust(
    results_ordered_df$pValue,
    method = "fdr"
)

write.csv(
  results_ordered_df,
  paste0(data.dir.single.cpg.pval, "MtSinai_single_cpg_pVal_df.csv"),
  row.names = FALSE
)

```

```{R}
results_ordered_df <- readr::read_csv(paste0(data.dir.single.cpg.pval, "MtSinai_single_cpg_pVal_df.csv"))
results_ordered_df
```


# Linear regression by regions median Methylation 

## Residuals control and coMethylated Regions

1. Take residuals
2. Find co-methylated regions using residuals

Input: 

- QNBMIQ_PCfiltered
- pheno_withNeuronProp_df

Output: 

- QNBMIQ_PCfiltered_mvalResiduals
- residuals_cometh_ls

```{R, eval = FALSE}
##### 1. Import datasets #######################################################
beta_mat <- readRDS(grep("QNBMIQ",dir(data.dir.pca,full.names = T),ignore.case = T,value = T)) 
pheno_df <- readRDS(dir(data.dir.neuron,full.names = T)) 

### Compute M values
mvalue_mat <- log2(beta_mat / (1 - beta_mat))

### Reorder samples based on pheno_df
mvalue_mat <- mvalue_mat[, pheno_df$sample]

identical(colnames(mvalue_mat),  pheno_df$sample)

### Take residuals
lmF <- function(mval){
  fitE <- lm(
    as.numeric(mval) ~ age.brain + sex + prop.neuron + as.character(slide), #add batch if rosmap
    data = pheno_df,
    na.action = na.exclude
  )
  residuals (fitE)
}
doParallel::registerDoParallel(cores = detectCores()/2)
resid <- plyr::adply(mvalue_mat,1,.fun = lmF,.progress = "time",.parallel = TRUE)
rownames(resid) <- resid[,1]
resid[,1] <- NULL
colnames(resid) <- colnames(mvalue_mat)
dim(resid)
# [1] 433656     22

dim(mvalue_mat)
# [1] 433656     22

### Save dataset
saveRDS(
  resid,
  paste0(data.dir.residuals, 
         "MtSinai_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)


##### 4. Find co-methylated regions ############################################

### Import datasets
mvalue_residuals_mat <- readRDS(
  paste0(data.dir.residuals, 
         "MtSinai_QNBMIQ_PCfiltered_mvalResiduals.RDS")
)

### Call in functions
library(coMethDMR)
library(BiocParallel)

probes.cluster.all <- coMethDMR::getPredefinedCluster(arrayType = "450k",
                                                      clusterType = "regions")

ncores <- 10
### Find co-methylated clusters
coMeth_ls <- CoMethAllRegions(
  dnam = mvalue_residuals_mat,      
  betaToM = FALSE,                   
  CpGs_ls = probes.cluster.all,
  arrayType = "450k",
  rDropThresh_num = 0.4,
  minPairwiseCorr = NULL,
  method = "spearman",             
  returnAllCpGs = TRUE,              
  output = "all",
  nCores_int = ncores,
  progressbar = FALSE
)

saveRDS(
  coMeth_ls,
  paste0(data.dir.residuals, 
         "MtSinai_residuals_cometh_input_ls.RDS")
)
```


## Linear regression by regions median Methylation 

1. Calculate medians by regions (found using the residuals) and sample using M-values
2. linear regression

Input: 

- QNBMIQ_PCfiltered,
- pheno_withNeuronProp_df
- residuals_cometh_input_ls

Output: 

- info_df
- mediansMval_df
- linear_df


### Calculate medians by cluster and sample

```{R, eval = FALSE}
### Import datasets
beta_mat <- readRDS(dir(data.dir.pca, pattern = "QNBMIQ", full.names = TRUE))

pheno_df <- readRDS(dir(data.dir.neuron,full.names = T)) 

mval_mat <- log2(beta_mat / (1 - beta_mat)) %>% as.matrix()
coMeth_ls <- readRDS(
  paste0(data.dir.residuals, 
         "MtSinai_residuals_cometh_input_ls.RDS")
)

### Create info dataset
input_cometh <- data.frame(
  inputRegion = coMeth_ls$inputRegion_chr,
  nCoMethRegion = coMeth_ls$nCoMethRegions_num,
  coMethRegion = names(coMeth_ls$coMeth_ls),
  nCpGs = unlist(lapply(coMeth_ls$coMeth_ls, length), use.names = FALSE),
  stringsAsFactors = FALSE
)

input_cometh_nodup <- input_cometh[
  !duplicated(input_cometh$coMethRegion),
  ]
colnames(input_cometh_nodup) <- c(
  paste0(cohort, "_inputRegion"),
  paste0(cohort, "_nCoMethRegion"),
  paste0(cohort, "_coMethRegion"),
  paste0(cohort, "_nCpGs")
)

saveRDS(
  input_cometh_nodup,
  paste0(data.dir.median, cohort, "_info_df.rds")
)

### Take median of probes in each cluster for each sample
filename <-  paste0(paste0(data.dir.median,cohort, "_mediansMval_df.rds"))

library(robustbase)
mval_mat <- mval_mat[rownames(mval_mat) %in% unlist(coMeth_ls$coMeth_ls),]
if(!file.exists(filename)){
  medianMval.df <- plyr::ldply(
    coMeth_ls$coMeth_ls[!duplicated(names(coMeth_ls$coMeth_ls))],
    function(probes){
      colMedians(mval_mat[as.character(probes),], na.rm = TRUE)
    },
    .progress = "time"
  )
  medianMval.df$.id <- NULL
  colnames(medianMval.df) <- colnames(mval_mat)
  saveRDS(medianMval.df, file = filename)
} else {
  medianMval.df <- readRDS(filename)
}
```

### Test all regions -- linear regressions

```{R, eval = TRUE}
### Import datasets
info_df <- readRDS(dir(data.dir.median,pattern = "info", full.names = TRUE))
mediansMval_df <- readRDS(dir(data.dir.median,pattern = "mediansMval", full.names = TRUE))
pheno_df <- readRDS(dir(data.dir.neuron, full.names = TRUE))

### Check variables before fitting model
pheno_df$Sample <- pheno_df$sample

identical(pheno_df$Sample, colnames(mediansMval_df))


pheno_df$sex <- as.factor(pheno_df$sex)
pheno_df$slide <- as.factor(pheno_df$slide)
# If rosmap cohort, don't forget batch effect

str(pheno_df)
```


```{R, eval = FALSE}
predictors_char <- "stage"
covariates_char <- c("age.brain", "sex", "prop.neuron", "slide")

devtools::source_gist("https://gist.github.com/tiagochst/d3a7b1639acf603916c315d23b1efb3e")

res_df <- TestAllRegions_noInfo(
  predictors_char = predictors_char,
  covariates_char = covariates_char,
  pheno_df = pheno_df,
  summarizedRegions_df = mediansMval_df
)

colnames(res_df) <- c(
  paste0(cohort, "_estimate"),
  paste0(cohort, "_se"),
  paste0(cohort, "_pVal"),
  paste0(cohort, "_fdr")
)

res_withInfo_df <- cbind(info_df, res_df)

saveRDS(
  res_withInfo_df,
  paste0(data.dir.median,cohort, "_linear_df.rds")
)
```

```{R}
file <- dir(data.dir.median,pattern = paste0(".*linear_df"),
            recursive = T,
            full.names = TRUE,
            ignore.case = T)
file
res_withInfo_df <- readRDS(file)
dim(res_withInfo_df)
res_withInfo_df
```

# Data final

```{R}
dir(path = data.dir,recursive = T,pattern = ".rda|.csv|.RDS")
```

# Session information
```{R}
devtools::session_info()
```
